A meta analysis of tournaments and an evaluation of performance in the Iterated Prisoner's Dilemma.

### 2019-09-22 paper/main.tex
paper draft

where \(T > R > P > S\) and \(2R > T + S\). The most common values used in
the literature, and in this paper, are $R=3, P=1, T=5, S=0$.

Since the computer tournaments of R. Axelrod in 1980s several academic papers
are published in the field regarding the performance of strategies in the
IPD. In the 80's following the strong performance of Tit For Tat
in both Axelrod's computer tournaments~\cite{Axelrod1980a, Axelrod1980b}, and
moreover in a series of evolutionary experiments~\cite{Axelrod1981}, the strategy
was thought as the most robust basic strategy in the iterated prisoner's
dilemma. However, the strategy's poor performance in environments with
noise~\cite{Bendor1991, Donninger1986, Molander1985, Hammerstein1984},
made room for other protagonists, such as, Nice and Forgiving~\cite{Bendor1991},
Pavlov~\cite{Nowak1993} and Generous Tit For Tat~\cite{Nowak1992}.

In 2004, the Anniversary Iterated Prisoner Dilemma Tournament was run with 233
entries. The winning strategies were based on a mechanism of teams. A team from
Southampton University took advantage of the fact that a participant was allowed
to submit multiple strategies. They submitted a total of 60 of strategies that
could recognised each other and colluded to increase one members
score~\cite{J.P.Delahaye1993Lp, J.P.Delahaye1995LIeP, A.Rogers2007Ctpw}.

Yet again, in 2012 another set of strategies was introduced as the dominant set
of strategies~\cite{Press2012}. These were called zero-determinant strategies,
and by forcing a linear relationship between the payoffs they can ensure that
they will never receive less than their opponents. However, in~\cite{Harper2017}
a tournament containing over 200 strategies, zero-determinant, was performed and
none of the zero-determinant strategies ranked in top spots. Instead, the top
ranked strategies were a set of evolved strategies based lookup tables, hidden
markov models and finite state automata.
%TODO reference the archetypes of the archetypes used the IPD?

@@

The different tournaments and the data generating process are covered in
Section~\ref{section:data_collection}. Section~\ref{section:top_performances},
covers the best performed strategies for each type of tournament and overall.
Section~\ref{section:evaluation_of_performance}, explores the traits which
contribute to good performance and finally in Section~\ref{section:conclusion}
the results are discussed and summarised.

@@

The process of generating data implemented in this manuscript is given by
Algorithm~\ref{algorithm:data_generation}. For each trial a random
size \(N\) is selected, and from the list of 186 strategies
in~\cite{axelrodproject}, a random list of \(N\) strategies is chosen. For the
given list of strategies a standard, a noisy, a probabilistic ending and a noisy
probabilistic ending tournament are performed and repeated \(k\) times.
The parameters for the tournaments as well as the number of repetitions are
selected once for each trial. The parameters and their respective minimum and
maximum values are given by Table~\ref{table:parameters_values}.

@@

The normalised rank is a measure which was manually included. The normalised
rank, denoted as $r$, is calculated as a strategy's rank divided by the
tournament's size ($N$). The normalised rank will be used in the next section to
evaluate the performance of strategies.

@@

This section evaluates the performance of 186 strategies which can be found in
the Appendix. The performance of each strategy will be evaluated for each type
of tournament independently. In Section~\ref{subsection:standard_tournament} the
strategies are evaluated on their performance in standard tournaments, in
Section~\ref{subsection:noisy_tournament} in noisy, in
Sections~\ref{subsection:probend_tournament} in probabilistic ending and in
\ref{subsection:noisy_probend_tournament} in noisy probabilistic ending.

Each strategy could have participated in multiple tournaments of the same type
(on average each participated in 5690 different tournaments). For example Tit
For Tat has participated in a total of 5569 standard tournaments. The strategy's
normalised rank distribution in different tournament is given in
Figure~\ref{fig:tit_for_tat_r_distribution}. As a result, of the multiple
entries of strategies their performance is evaluated based on the median
normalised rank denoted as \(\bar{r}\). A value of \(\bar{r} = 0\) corresponds
to a strategy winning the tournament where a a value of \(\bar{r} = 1\)
corresponds to the strategy coming last.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.8\textwidth]{../images/tit_for_tat_r_distributions.pdf}
    \caption{Tit For Tat's $r$ distribution in tournaments.}
    \label{fig:tit_for_tat_r_distribution}
\end{figure}

Following the analysis of each tournament type, in Section~\ref{subsrction:overall}
the results are merged and the strategies are evaluated in their overall performances
over all simulated tournaments of this work.

\subsection{Top ranked strategies in: Standard tournaments}\label{subsection:standard_tournament}

The results presented in this section are based on $12,285$ different standard
tournaments.
The top 15 performances in standard tournaments are given in
Table~\ref{table:std_results}. Ten of these are strategies introduced
in~\cite{Harper2017}. These have been trained using reinforcement learning
algorithms and are based on finite state automata (FSM), hidden markov models
(HMM), artificial neural networks (ANN), lookup tables (LookerUp) and stochastic
lookup tables (Gambler). Specifically, the have been trained against the
strategy list of~\cite{axelrodproject} in standard tournaments. Thus their
performance is to be expected. DoubleCrosser, and Fool Me Once, are not from
the literature but from~\cite{axelrodproject}. DoubleCrosser is a strategy that
makes use of the length of the match because is set to defect on the last two
rounds. The strategy is expected to not perform as well in probabilistic ending
tournaments. Finally, Winner 12~\cite{Mathieu2017} and DBS are both from the the
literature. DBS~\cite{Au2006} is strategy specifically designed for noisy
environments, however, it ranks highly only in standard ones.

\begin{table}[!htbp]
    \centering
    \resizebox{.28\textwidth}{!}{
    \input{performance_standard_table.tex}}
    \caption{Standard top performances}\label{table:std_results}
\end{table}

Figure~\ref{fig:std_results} gives the distributions of $\bar{r}$ for the top
ranked strategies. The distributions are skewed towards zero and the highest
median is at 0.075. This indicates that the top ranked strategies are dominating
strategies in standard tournaments. They are very likely to perform well in
standard tournament despite the number of opponents, the opponents, the turns
etc. This does not hold for all the tournament types as it will be discussed
in the following sections. The next section present the results of noisy
tournaments.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/performance_standard.pdf}
    \caption{$\bar{r}$ distributions of top 15 strategies.}\label{fig:std_results}
\end{figure}

\subsection{Top ranked strategies in: Noisy tournaments}\label{subsection:noisy_tournament}

Similarly to Section~\ref{subsection:standard_tournament} the results presented
here are based on $12,285$ different noisy tournaments. The strategies have been
ranked based on their \(\bar{r}\) and the top 15 are given by
Table~\ref{table:noisy_results}. The distributions of their corresponding
\(\bar{r}\) is also given in Figure~\ref{fig:noisy_results}.

The top strategies include the two strategies, Tit For 2
Tats~\cite{Axelrod1980b} and Hard Tit For 2 Tats~\cite{Stewart2012}, which are
strategies that will defect only after the have received two defections from the
opponent. The Retaliate strategies are set of strategies
from~\cite{axelrodproject} that start by cooperating but will retaliate once the
opponent's wins and defections surpass a curtain threshold.
ShortMem~\cite{Carvalho2013}, Grumpy, $e$ and $\phi$ are strategies that make
decisions based on the cooperations to defections ratio. In $5^{\text{th}}$ and
$6^{\text{th}}$ place are the strategies Cycler Hunter and  Risky QLearn. Cycler
Hunter tries to extort strategies that play cyclically and Risky QLearn uses a Q
learning algorithm. Notably, a deterministic and one of the most simple
strategies in game is ranked $3^{\text{rd}}$. That is Cooperator, a strategy
that just cooperates.

\begin{table}[!htbp]
    \centering
    \resizebox{.25\textwidth}{!}{
    \input{performance_noise_table.tex}}
    \caption{Noisy top performances}\label{table:noisy_results}
\end{table}

@@

\subsection{Top ranked strategies: Probabilistic ending tournaments}\label{subsection:probend_tournament}

In this section the performances in probabilistic ending tournaments are
discussed. The 15 top ranked strategies are given by
Table~\ref{table:prob_end_results} and Figure~\ref{fig:probend_results} gives
their respective $\bar{r}$ distributions.
Fortress 3, Fortress 4 (both introduced in~\cite{Ashlock2006}),
Raider~\cite{Ashlock2014} and Solution B1~\cite{Ashlock2014} are strategies
based on finite state automata introduced by Daniel and Wendy Ashlock. These
strategies have been evolved using reinformed learning, however, there were
trained to maximise their payoffs in tournaments with fixed turns (150
specifically) and not in probabilistic ending ones.

In probabilistic ending tournaments it appears that the top ranks are mostly
occupied by defecting strategies which include Better and Better, Gradual
Killer, Hard Prober (all from ~\cite{prison}), Bully (Reverse Tit For
Tat)~\cite{Nachbar1992} and Defector. Thus, it's surprisingly that EasyGo and
Fool Me Forever are ranked $14^{\text{th}}$ and $15^{\text{th}}$. These
strategies are actually the same; they will defect until their opponent defect,
then they will cooperate until the end. Both strategies have repeatedly ranked highly
as shown in Figure~\ref{fig:probend_results} and there are cases for which they
were the winners of tournaments.

\begin{table}[!htbp]
    \centering
    \resizebox{.25\textwidth}{!}{
    \input{performance_probend_table.tex}}
    \caption{Probabilistic ending top performances}\label{table:prob_end_results}
\end{table}

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.8\textwidth]{../images/performance_probend.pdf}
    \caption{\(\bar{r}\) distributions for best performed strategies in probabilistic ending tournaments.}
    \label{fig:probend_results}
\end{figure}

\subsection{Noisy \& probabilistic ending tournaments}\label{subsection:noisy_probend_tournament}

This section presents the results in tournaments that have both noise and an
unspecified number of turns. Several of the top ranked have been covered before
because they were highly ranked in noisy tournaments as well,
Table~\ref{table:noisy_prob_end_results}. However, strategies from the
top ranks of probabilistic ending tournaments did not rank highly here.

The Retaliates, $\phi, e$ and Anti Tit For Tat behaviour strategies appear to
perform well in noisy environments. So these strategies performed well in this
setting as well even though now the turns are not specified. The top ranked
strategy is Alternator a strategy that alternate between cooperation and
defection. Hopeless~\cite{Van2015} is a strategy that will only defect if and
only if a mutual cooperation and the last three places are occupied by
strategies based on a Q learning algorithm.

\begin{table}[!htbp]
    \centering
    \resizebox{.25\textwidth}{!}{
    \input{performance_probend_noise_table.tex}}
    \caption{Noisy and probabilistic ending top performances}\label{table:noisy_prob_end_results}
\end{table}

The three Q learning strategies are the only ones that have bimodal
distributions of normalised ranks similar to those of
Section~\ref{subsection:noisy_tournament}. In comparison, the rest of the
distributions are skewed towards 0.3 to 0.4
Figure~\ref{fig:noisy_probend_results}.

@@

As yet, the performances of 186 different strategies have been evaluated in
four different tournaments. In standard tournaments the top ranked spots were
dominated by complex strategies that have been trained using reinforcement learning
techniques in a standard tournament. These strategies dominated most
the tournaments that have been involved in. In noisy tournaments the top ranked
strategies were all different to those of standard tournaments. Most of the
top ranked strategies have been strategies that were trying to keep the defections
and cooperations to a given ratio. These strategies had won most the tournaments
in which they participated, however, they also ranked last in several tournaments.
Once again in probabilistic ending tournaments the top ranked strategies were
different to those before. More defecting strategies occupied the top ranks
as well as finite state automata that have been introduced all by the same
authors. Finally, in noisy probabilistic ending tournaments it was the only
set of tournaments were the highly ranked strategies have been covered before,
more specifically, in noisy tournaments. However, in these tournaments the means
and medians of the distributions were close to 0.4. Indicating that on average
these strategies manage to beat only 60\% of the opponents.

In the following section the performance of the strategies is evaluated again
but this time the results of the tournament types are included in the
analysis.

\subsection{Top performance in data set}\label{subsrction:overall}

The data set described in Section~\ref{section:data_collection} contains results
from 49,140 tournaments and the performances in this section are 
based on these tournaments. The top ranked 15 strategies are given in
Table~\ref{table:overall_results}. It includes several strategies that have
ranked in the top spots in the separate tournament analysis.
The top spaces are overtaken by the Retaliate strategies, followed by BackStabber
and DoubleCrosser. DoubleCrosser is just an extension of BackStabber.
Nice Meta Winner and NMWE Memory One are strategies based on teams of strategies
and Stein and Rapoport and Grudger are strategies from Axelrod's original tournament
where they came 6th and 7th respectively. Forgetful Fool Me Once is a similar
strategy to Grudger but sometimes forgets and cooperates. Finally, both
PSO Gambler and Evolved HMM 5 are strategies introduced in~\cite{Harper2017}.

@@

The top ranked strategies include a good mixture of strategies. PSO Gambler and
Evolved HMM 5 are very complex strategies compared to Stein and Rapoport and
Grudger that are strategies manually designed. There are strategies based on
teams and the Retaliate strategies which have illustrated to do well in noise.
The Retaliates' distributions are very similar to each other,
Figure~\ref{fig:overall_results}, and different compared to the rest of the
distributions.

@@

However, are these strategies the answer as to how to play the IPD. The distributions
of their ranks are given for all the tournaments individually. Figure illustrates
that there are settings that these strategies perform well, that they perform
badly and settings that they perform average. Thus the conclusion is that there
is not a single dominant strategy in the IPD.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.8\textwidth]{../images/performance.pdf}
    \caption{\(\bar{r}\) distributions for best performed strategies in data set.}
    \label{fig:overall_results}
\end{figure}

In this section the top ranked performances were presented. This was not done in
a single tournament of a given list of opponents, as many in the literature, but
over a number of tournaments where the list and size could vary. Moreover, four
different type of tournaments were performed in this work. Even with the large
number of tournaments there some think that is evident, there is not a single
best strategy. Though some strategies seem to be repeated, there are other
tournament types that they are not. The top ranks are occupied but almost
different strategies every time. And in the top spots the strategies behaviour
differ. There are simple and complex, with small memory and large, deterministic
and stochastic. Is there a single dominant strategy in the iterated prisoner's
dilemma? We argue not.

The aim of the next section is to understand which are the factors that made
these strategies successful. In each setting separately but also overall.

@@

This section explores the factors that contributed to the successful performance
of strategies described in Section~\ref{section:top_performances}.
The factors
evaluated in this work are given by Table~\ref{table:manual_measures}. Several
of these factors include measures regarding the strategies and it's behaviour
however others are fixed from the tournament or even the type of tournament. The
measures from Table 1 are also included.

This is achieved by
separating the strategies performances into successful ones and non. The performances
are separated on the normalised rank and the median score using the clustering
algorithm \(k-\)means~\cite{Arthur2007}. The number of clusters is not deterministic
but the decision is based on the silhouette coefficients~\cite{Rousseeuw1987}.

Consider the performances in the standard tournaments.
Figure~\ref{fig:standard_clusters} gives the normalised rank against the median
score for three trials of the clustering algorithm with number of clusters being
2, 3 and 4. A number of 2 clusters gives the highest silhouette coefficient (0.66)
and so it's selected for the standard tournaments. More specifically,
being in cluster 1 corresponds to a low performance whereas being in cluster
2 corresponds to a high performance. These are based on the median normalised
rank and median score of each cluster as shown in Table~\ref{table:clusters}.

@@

\begin{table}[!htbp]
    \centering
    \resizebox{.25\textwidth}{!}{
    \begin{tabular}{lcc}
    \toprule
    Cluster                &   $\bar{r}$ & median score \\
    \midrule
    1 & 0.769 & 2.065 \\
    2 & 0.263 & 2.639\\
    \bottomrule
    \end{tabular}}
    \caption{Median normalised rank and median score of each cluster in standard tournaments.}
    \label{table:clusters}
\end{table}

>>>>

where \(T > R > P > S\) and \(2R > T + S\). The most common values used in
the literature~\cite{Axelrod1981}, and in this paper, are $R=3, P=1, T=5, S=0$.

Since the computer tournaments of R. Axelrod in 1980s several academic papers
are published in the field regarding the performance of strategies in the IPD.
In the 80's following the strong performance of Tit For Tat in both Axelrod's
computer tournaments~\cite{Axelrod1980a, Axelrod1980b}, and moreover in a series
of evolutionary experiments~\cite{Axelrod1981}, the strategy was thought as the
most robust basic strategy in the IPD. However, the strategy was shown to
perform poorly in environments with noise~\cite{Bendor1991, Donninger1986,
Molander1985, Hammerstein1984}. More robust strategy in such setting were
introduced and were the new protagonists, such as, Nice and
Forgiving~\cite{Bendor1991}, Pavlov~\cite{Nowak1993} and Generous Tit For
Tat~\cite{Nowak1992}.

The $20^{\text{th}}$ Anniversary Iterated Prisoner Dilemma Tournament took place
in 2004 with 233 entries. The winning strategies were based on a mechanism of
teams. A team from Southampton University took advantage of the fact that a
participant was allowed to submit multiple strategies. They submitted a total of
60 of strategies that could recognised each other and colluded to increase one
members score~\cite{J.P.Delahaye1993Lp, J.P.Delahaye1995LIeP, A.Rogers2007Ctpw}.
Yet again, in 2012 another set of strategies was introduced as the dominant set
of strategies~\cite{Press2012}. These were called zero-determinant strategies,
and by forcing a linear relationship between the payoffs they can ensure that
they will never receive less than their opponents. However, in~\cite{Harper2017}
a tournament containing over 200 strategies, zero-determinant, was performed and
none of the zero-determinant strategies ranked in top spots. Instead, the top
ranked strategies were a set of evolved strategies based lookup
tables~\cite{Axelrod1987}, hidden markov models~\cite{Harper2017} and finite
state automata~\cite{Miller1996}.
%TODO reference the archetypes of the archetypes used the IPD? Did IPD for now.

@@

The different tournaments and the data generating process are covered in
Section~\ref{section:data_collection}. Section~\ref{section:top_performances},
covers the best performed strategies for each type of tournament and overall.
Section~\ref{section:evaluation_of_performance}, explores the traits which
contribute to good performance and finally in Section~\ref{section:discussion}
the results are discussed and concluded in Section~\ref{section:conclusion}.

@@

The process of generating data implemented in this manuscript is given by
Algorithm~\ref{algorithm:data_generation}. For each trial a random
size \(N\) is selected, and from the list of 186 strategies
in~\cite{axelrodproject}, a random list of \(N\) strategies is chosen. The 186
strategies used here are given in the Appendix~\ref{app:list_of_players}. For the
given list of strategies a standard, a noisy, a probabilistic ending and a noisy
probabilistic ending tournament are performed and repeated \(k\) times.
The parameters for the tournaments as well as the number of repetitions are
selected once for each trial. The parameters and their respective minimum and
maximum values are given by Table~\ref{table:parameters_values}.

@@

The \textbf{normalised rank} is a measure which was manually included. The normalised
rank, denoted as $r$, is calculated as a strategy's rank divided by the
tournament's size ($N$). The normalised rank will be used in the next section to
evaluate the performance of strategies.

@@

This section evaluates the performance of 186 strategies. The performance of
each strategy will be evaluated for each type of tournament independently,
followed by an evaluation of their performance over all the simulated
tournaments of this work. Each strategy could have participated in multiple
tournaments of the same type (on average each participated in 5690 different
tournaments). For example Tit For Tat has participated in a total of 5569
tournaments of each type. The strategy's normalised rank distribution in these
is given in Figure~\ref{fig:tit_for_tat_r_distribution}. As a result, of the
multiple entries of strategies their performance is evaluated based on the
\textbf{median normalised rank} denoted as \(\bar{r}\). A value of \(\bar{r} =
0\) corresponds to a strategy winning the tournament where a a value of
\(\bar{r} = 1\) corresponds to the strategy coming last.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.8\textwidth]{../images/tit_for_tat_r_distributions.pdf}
    \caption{Tit For Tat's $r$ distribution in tournaments. The best performance
    of the strategy has been in standard tournaments where it achieved a normalised
    rank of 0.339.}
    \label{fig:tit_for_tat_r_distribution}
\end{figure}

The result of each tournament type are based on $12,285$ trials. The strategies
are ranked based on the median normalised rank they achieved and the top 15
strategies of each tournament are given in Table~\ref{table:top_performances}.

\newcolumntype{g}{>{\columncolor{Gray}}l}
\begin{table}[!htbp]
    \begin{center}
    \resizebox{.9\textwidth}{!}{
        \input{top_perfomances.tex}
    }
\end{center}
\caption{Top performances for each tournament type based on $\bar{r}$.}
\label{table:top_performances}
\end{table}

In standard tournaments 10 of the 15 top strategies are strategies introduced
in~\cite{Harper2017}. These have been trained using reinforcement learning
algorithms and are based on finite state automata (FSM), hidden markov models
(HMM), artificial neural networks (ANN), lookup tables (LookerUp) and stochastic
lookup tables (Gambler). Specifically, the have been trained against the
strategy list of~\cite{axelrodproject} in standard tournaments. Thus their
performance is to be expected. DoubleCrosser, and Fool Me Once, are not from the
literature but from~\cite{axelrodproject}. DoubleCrosser is a strategy that
makes use of the length of the match because is set to defect on the last two
rounds. The strategy is expected to not perform as well in probabilistic ending
tournaments. Finally, Winner 12~\cite{mathieu2017} and DBS are both from the the
literature. DBS~\cite{Au2006} is strategy specifically designed for noisy
environments, however, it ranks highly only in standard ones.

Figure~\ref{fig:std_results} gives the distributions of $\bar{r}$ for the top
ranked strategies. The distributions are skewed towards zero and the highest
median is at 0.075. This indicates that the top ranked strategies are dominating
strategies in standard tournaments. They are very likely to perform well in
standard tournament despite the number of opponents, the opponents, the turns
etc. This does not hold for all the tournament types as it will be discussed
in the later parts.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/performance_standard.pdf}
    \caption{$\bar{r}$ distributions of top 15 strategies in standard tournaments.}\label{fig:std_results}
\end{figure}

The top strategies in noisy tournaments include the two strategies, Tit For 2
Tats~\cite{Axelrod1980b} and Hard Tit For 2 Tats~\cite{Stewart2012}, which are
strategies that will defect only after the have received two defections from the
opponent. The Retaliate strategies are set of strategies
from~\cite{axelrodproject} that start by cooperating but will retaliate once the
opponent's wins and defections surpass a curtain threshold.
ShortMem~\cite{Carvalho2013}, Grumpy, $e$ and $\phi$ are strategies that make
decisions based on the cooperations to defections ratio. In $5^{\text{th}}$ and
$6^{\text{th}}$ place are the strategies Cycler Hunter and  Risky QLearn. Cycler
Hunter tries to extort strategies that play cyclically and Risky QLearn uses a Q
learning algorithm. Notably, a deterministic and one of the most simple
strategies in game is ranked $3^{\text{rd}}$. That is Cooperator, a strategy
that just cooperates.

@@

The 15 top ranked strategies in probabilistic ending tournaments include
Fortress 3, Fortress 4 (both introduced in~\cite{Ashlock2006}),
Raider~\cite{Ashlock2014} and Solution B1~\cite{Ashlock2014} which are strategies
based on finite state automata introduced by Daniel and Wendy Ashlock. These
strategies have been evolved using reinformed learning, however, there were
trained to maximise their payoffs in tournaments with fixed turns (150
specifically) and not in probabilistic ending ones.
In probabilistic ending tournaments it appears that the top ranks are mostly
occupied by defecting strategies which include Better and Better, Gradual
Killer, Hard Prober (all from ~\cite{prison}), Bully (Reverse Tit For
Tat)~\cite{Nachbar1992} and Defector. Thus, it's surprisingly that EasyGo and
Fool Me Forever are ranked $14^{\text{th}}$ and $15^{\text{th}}$. These
strategies are actually the same; they will defect until their opponent defect,
then they will cooperate until the end. Both strategies have repeatedly ranked highly
as shown in Figure~\ref{fig:probend_results} and there are cases for which they
were the winners of the tournament.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=.7\textwidth]{../images/performance_probend.pdf}
    \caption{\(\bar{r}\) distributions for best performed strategies in probabilistic ending tournaments.}
    \label{fig:probend_results}
\end{figure}

In tournaments that have both noise and an unspecified number of turns several
of the top ranked strategies are strategies that were highly ranked in
noisy tournaments as well. However, strategies from the top ranks of
probabilistic ending tournaments did not rank highly here. The Retaliate set,
$\phi, e$ and Anti Tit For Tat behaviour strategies appear to perform well in
noisy environments. So these strategies performed well in this setting as well
even though now the turns are not specified.
The top ranked strategy is Alternator a strategy that alternate between
cooperation and defection. Hopeless~\cite{Van2015} is a strategy that will only
defect if and only if a mutual cooperation and the last three places are
occupied by strategies based on a Q learning algorithm. The three Q learning
strategies are the only ones that have bimodal distributions of normalised
ranks. In comparison, the rest of the distributions are skewed towards
0.4, Figure~\ref{fig:noisy_probend_results}.

@@

@@

So far the performances have been evaluated separately for each tournament
type. The merged data set, which was described in Section~\ref{section:data_collection},
contains a total of 49,140 result summaries. The 15 top ranked strategies
overall are given in Table~\ref{table:overall_results}. The top ranks
include strategies that have been mentioned before due to their performance
in specific tournaments. The top ranks are overtaken by the set of Retaliate
strategies followed by BackStabber and DoubleCrosser. DoubleCrosser performed
well in standard tournaments and the strategy is just an extension of BackStabber.
The two strategies Nice Meta Winner and NMWE Memory One are strategies based on teams.
PSO Gambler and Evolved HMM 5 are trained strategies introduced in~\cite{Harper2017}
and Stein and Rapoport and Grudger are strategies from Axelrod's
original tournament where they came 6th and 7th respectively. Forgetful Fool Me
Once is based on the same approach as Grudger.

Figure~\ref{fig:overall_results} gives the normalised rank distributions
of these strategies. It is evident that the Retaliate strategies performances
are very similar and their distributions are different in comparison with the
rest of the top ranked strategies in the setting.

@@

This section presented the top ranked performances.

\begin{itemize}
    \item In standard tournaments the top ranked spots were dominated by complex
    strategies that have been trained using reinforcement learning techniques.
    These strategies dominated most the tournaments that have been involved in.
    \item In noisy tournaments the top ranked strategies were all different to those of standard
    tournaments. Most of the top ranked strategies were basing their decisions
    on the he defection to cooperation ratio. These
    strategies had won most the tournaments in which they participated, however,
    they also ranked last in several tournaments.
    \item Once again in probabilistic ending tournaments the top ranked
    strategies were different to those before. More defecting strategies
    occupied the top ranks in this setting, as well as finite state automata
    that have been introduced all by the same authors.
    \item Moreover, in noisy probabilistic ending tournaments it was the only
    set of tournaments were the highly ranked strategies have been covered
    before, more specifically, in noisy tournaments. In these tournaments the
    means and medians of the distributions were close to 0.4. Indicating that on
    average these strategies managed to consistently perform on the top 40\% of
    the strategies.
    \item Finally, using all the tournaments results, despite their type, the top
    ranked strategies were a mixture of behaviours that do well in standard
    tournaments and tournaments with noise. The top ranks also included two
    strategies that are based on teams.
\end{itemize}

In this section the winning strategies in a series of IPD tournaments of
different settings were presented. Though there have been instances of
strategies that do well in more than one setting, most of the top ranks were
occupied by different strategies each time. The top ranks have been occupied
by strategies with small memory size, with infinite memory size, sophisticated
and deterministic strategies. Even two of the most basic strategic rules,
Cooperator and Defector, have been ranked in the top 15 places.

This indicates that in an IPD tournament a winning strategy exist but only for
the given environment, however, there is no dominant strategy in the IPD that
can do well in any tournament where the settings can vary. The aim of the next
section is to understand which are the factors that made these strategies
successful. In each setting separately but also overall.

@@

The aim of this section is to explore the factors that contribute to a
strategy's successful performance. The factors explored are measures regarding a
strategy's behaviour but also measures regarding the tournaments the strategies
competed in. More specifically the factors that are included in the analysis are
given by Table~\ref{table:manual_measures} and
Table~\ref{table:parameters_values}.

Axelrod-Python makes use of classifiers to classify strategies according to
various dimensions. The classifiers considered in this work are the stochastic
classifier, the make use of game and length classifiers. These determine whether
a strategy is stochastic or deterministic, whether it makes use of the number of
turns or the game's payoffs. The SSE is a measure extortionate
behaviour introduced in~\cite{Knight2019}. A value of 1 indicates no
extortionate behaviour at all whereas a value of 0 indicates that a strategy is
always trying to extortion the opponent. The rest of the factors considered
are the $CC$ to $C$, $CD$ to $C$, $DC$ to $C$, and $DD$ to $C$ rates as well as
cooperating ratio of a strategy. The minimum, maximum, medium and median
cooperating ratios of each tournament are also included, and finally the number of
turns, the size and the probabilities of noise or the game ending.

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[h]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{gcgcgc}
    \toprule
    measure & measure explanation &  source & value type & min value & max value \\
    \midrule
stochastic  &  If a strategy is stochastic & strategy classifier from~\cite{axelrodproject} & boolean  & False &  True \\
makes use of game &  If a strategy makes used of the game information & strategy classifier from~\cite{axelrodproject} & boolean  & False &  True \\
makes use of length &  If a strategy makes used of the number of turns & strategy classifier from~\cite{axelrodproject} & boolean  & False &  True \\
SSE & A measure of how far a strategy is from extortionate behaviour & method described in~\cite{Knight2019} & float & 0 & 1 \\
max cooperating rate $(C_{\text{max}})$  & The biggest cooperating rate in the tournament & result summary & float & 0 & 1 \\
min cooperating rate $(C_{\text{min}})$ & The smallest cooperating rate in the tournament & result summary & float & 0 & 1 \\
median cooperating rate $(C_{\text{median}})$ & The median cooperating rate in the tournament & result summary & float & 0 & 1 \\
mean cooperating rate $(C_{\text{mean}})$ & The mean cooperating rate in the tournament & result summary & float & 0 & 1 \\
$C_r$ / $C_{\text{max}}$ & A strategy's cooperating rate divided by the maximum & manually & float & 0 & 1 \\
$C_r$ / $C_{\text{min}}$ & A strategy's cooperating rate divided by the minimum & manually & float & 0 & 1 \\
$C_r$ / $C_{\text{median}}$ & A strategy's cooperating rate divided by the median & manually & float & 0 & 1 \\
$C_r$ / $C_{\text{mean}}$ & A strategy's cooperating rate divided by the mean & manually & float & 0 & 1 \\
$C_r$ & The cooperating ratio of a strategy & result summary & float & 0 & 1 \\
$CC$ to $C$ rate & The probability a strategy will cooperate after a mutual cooperation & result summary & float & 0 & 1 \\
$CD$ to $C$ rate & The probability a strategy will cooperate after being betrayed by the opponent & result summary & float & 0 & 1 \\
$DC$ to $C$ rate & The probability a strategy will cooperate after betraying the opponent & result summary & float & 0 & 1 \\
$DD$ to $C$ rate & The probability a strategy will cooperate after a mutual defection & result summary & float & 0 & 1 \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{Manually calculated/retrieved measures.}
    \label{table:manual_measures}
\end{table}

The effect of these factors on a strategy's success is evaluated based on the
correlation coefficient between the factors, the normalised rank and the median
score. The correlation coefficients are given by Table~\ref{table:correlations}.
Note that the correlation for the classifiers is not included because they are
binary variables and they will evaluated by a different method.

% The
% correlations of each factor against each factor, the normalised rank and the
% median are also given in the Appendix.

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[!htbp]
    \begin{center}
    \resizebox{.8\textwidth}{!}{
        \input{correlations_table.tex}
    }
\end{center}
\caption{Correlations table.}\label{table:correlations}
\end{table}

In standard tournaments the measures  $CC$ to $C$, $C_r$, $C_r / C_{\text{max}}$
and the cooperating ratio compared to the median and the mean have a moderate
negative effect on the normalised rank and a moderate positive on the median
score. The SSE error and the $DD$ to $C$ have the opposite effects. Thus, in
standard tournaments behaving cooperative corresponds to a more successful
performance. However, even though being cooperative pays of that's not true
against defective strategies. Cooperating after a mutual defection lowers a
strategy's success. Figure\ref{fig:rates_of_winners_in_standard_tournaments}
confirms these. The winner of standard tournaments almost always cooperated
after a mutual cooperation but the 50\% only cooperated with a probability
of 0.2 after a mutual defection.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/rates_of_winners_in_standard_tournaments.pdf}
    \caption{Winners distributions
    in standard tournaments for $CC$ to $C$ and $DD$ to $C$.}\label{fig:rates_of_winners_in_standard_tournaments}
\end{figure}

Compared to standard tournaments in both noisy and in probabilistic ending
tournaments the higher the rates of cooperation the lower a strategy's success
and median score. Moreover, a strategy would want to cooperate less than both
the mean and median cooperator in such settings. In probabilistic ending
tournaments the correlations coefficients have a larger values, indicating a
stronger effect. Thus a strategy will be punished more by it's cooperative
behaviour in such environments. The distributions of the $C_r$ of the winners in
both tournaments is given by Figure~\ref{fig:c_r_distributions}. It confirms
that the winners in these two settings were more defective strategies, mainly
in probabilistic ending ones.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/c_r_winners_tournaments.pdf}
    \caption{$C_r$ distributions.}\label{fig:c_r_distributions}
\end{figure}

In noisy probabilistic ending tournaments and in the overall tournaments the
only factors that had a moderate affect are $C_r/C_{\text{mean}},
C_r/C_{\text{max}}$ and $C_r$. In such environments cooperative behaviour
appears to be punished by not as much as in noisy and probabilistic ending
tournaments.

To further evaluate the effect of factors on performances a random forest
classification~\cite{breiman2001} is applied. Initially, the performances are
clustered based on their normalised rank and the median score by a \(k-\)means
algorithm~\cite{Arthur2007}. The number of clusters are not deterministically
chosen but are based on the silhouette coefficients~\cite{Rousseeuw1987}.
Consider the case of standard tournaments, the chosen number of clusters is
2 and Figure~\ref{fig:standard_clusters} illustrates the trials of clustering the
performances in 2, 3 and 4 clusters respectively. The number of chosen clusters
for each type and overall are given in Table~\ref{table:number_of_clusters}.

@@

\begin{table}[!htbp]
    \begin{center}
        \resizebox{.5\textwidth}{!}{
        \begin{tabular}{lcccc}
    \toprule
    type & number of clusters &  silhouette coefficient\\
    \midrule
    standard                   & 2 & 0.648 \\
    noisy                      & 3 & 0.493 \\
    probabilistic ending       & 2 & 0.680 \\
    noisy probabilistic ending & 4 & 0.415\\
    overall                    & 3 & 0.443\\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{Number of clusters for each type and overall and the respective silhouette coefficients.}
    \label{table:number_of_clusters}
\end{table}
